%insert blank page
\newpage\null\pagestyle{plain}\newpage

% header with section title
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\sectionmark}[1]{\markright{#1}} %no number
\fancyhead[le,ro]{\nouppercase{\rightmark}}
\fancyfoot[le,ro]{\thepage}
\renewcommand{\headrulewidth}{.4pt}
\renewcommand{\footrulewidth}{.4pt}

%!TEX root = ../dissertation.tex
\begin{savequote}[82mm]
If the human brain were so simple that we could understand it, we would be so simple that we couldn’t.
\qauthor{Emerson M. Pugh}
\end{savequote}

\chapter{Discussion}
The general objective of this dissertation was to facilitate an understanding of the complex processes surrounding transcriptional interactions in mammalian cells. The multi-leveled relationships of only two interacting molecule species make analysis and interpretation inaccessible without the help of software, and even the output of software analyses can be overwhelming in the multitude of genes that are involved. Additionally, while this dissertation started with the objective of illuminating cholinergic processes exclusively in neurons, it naturally gravitated towards immunology in all of the different foci: the studied degenerative and non-degenerative psychiatric diseases, as well as stroke, all have significant immunological components. Arguably, immune cells and the central nervous system are those two mammalian tissues that offer the greatest challenges to the life sciences in terms of complexity. 

Historically, the areas of immunology and neuroscience research have little in common, and translational advances have been few. However, as Robert Dantzer illustrates in his recent review,\cite{Dantzer2018} the two disciplines have much to learn from each other, and bringing them closer together is all but necessary. Not by coincidence, the description of brain-to-immune-signalling in that review is predominated by cholinergic implications in immunity; and the best-studied immune-to-brain-signalling molecules are the endogenous pyrogens IL-1$\upbeta$, IL-6, TNF-$\upalpha$, and IFN-$\upalpha$, which in this dissertation also are frequently implicated. However, current research barely scratches the surface of neuro-immune communication; currently, interactions are mainly studied at the level of cell-to-cell or pro"-tein-to-pro"-tein (also including transcriptional processes induced by proteins). An enormous amount of transcriptional interactions are still almost completely in the dark, including but not limited to small RNA regulatory processes. The main challenge in adding an additional regulatory layer onto an already complex subject such as neuro-immune communication is the resulting exponential increase in complexity.

This brings us back to the initial objective of this dissertation, the facilitation of an understanding of transcriptional interactions. Since the data generated by modern life-science technologies is not comprehensible to humans in its raw form, and even after statistical analyses often remains overwhelming, dimensionality reduction is a logical step to further the comprehension of the science by the scientist. Indeed, most approaches described in this dissertation result in reduction of dimensionality, and a common train of thought behind the distinct analysis steps undertaken was governed by the idea of a »smart« dimensionality reduction, as opposed to, for instance, exclusively looking at the miRNA$\to$gene relationship with the lowest p-value.

To this end, I had to develop a computational basis for the assessment of transcriptional interactions in a manner that is practicable in day-to-day research, i.e., that can generate results for these complex interactions in a matter of seconds to hours. I also had the fortune to be able to apply these methods to a range of relevant biological data, including the ones discussed in-depth in this dissertation: the cellular model of human male and female cholinergic neurons, and the blood of stroke patients. All undertaken analyses are subject to a wide variety of limitations, the most important of which will be discussed in the following.

For the sake of clarity, the discussion will be split into parts: first, the methodological and technical aspects; second, the bio-mechanistic perspective and basic molecular biology implications; and third, the physiologic, pathologic, and medical/therapeutic inferences.

\input{chapters/101_methods}
\input{chapters/102_mechanisms}
\input{chapters/103_therapy}